---
title: "Metabolite ILE Analysis"
author: "Andrew Quinn"
date: "1/28/2022"
md_document:
    variant: markdown_github
---
# Analysis of 13C Isotope labelling of gut metabolites in bees fed 100% U-13C glucose before switching to naturally abundant glucose at time zero.  

### These are the required package to reproduce this analysis.
```{r, echo = T, eval=T, message=F, warning=F, results=F}

### env setup

#required_packages = c( "data.table", "readxl",  "tidyverse", "plyr", "writexl", "tibble",  "wesanderson", "RColorBrewer" , "ggh4x", "purrr")
#need_install = required_packages[!(required_packages) %in% installed.packages()]

#if (length(need_install) > 0){
#  install.packages((need_install))
#}


#lapply(required_packages, require, character.only=T)

#dir_script = dirname(rstudioapi::getSourceEditorContext()$path)

#setwd(dir_script)
```

##This code combines processed 12C/13C isotopic abundances in gut metabolites from experimental replicates across time and condition
### Re-baseline time to begin at 13C --> 12C transition
```{r, echo = T, eval=T, message=F, warning=F, results=F}
library(data.table)
library(readxl)
library(plyr)
library(writexl)
library(tibble)
library(wesanderson)
library(RColorBrewer)
#library(ggh4x)
library(purrr)
library(tidyverse)
```

```{r, warning=F}
xlsx_data <- "Metabolite_ILE_results.xlsx"

excel_sheets(path = xlsx_data)

tab_names <- excel_sheets(path = xlsx_data)

list_all <- lapply(tab_names, function(x) {     
  as.data.frame(read_excel("Metabolite_ILE_results.xlsx", sheet = x)) } )

full_data <- do.call(rbind.fill, list_all)

full_data$MW <- factor(full_data$MW)
full_data$MI <- factor(full_data$MI)

full_data_1 <- full_data %>%
  mutate(Time = case_when(Time == 24 ~ 0,
                          Time == 32 ~ 8,
                          Time == 48 ~ 24,
                          Time == 56 ~ 32,
                          Time == 72 ~ 48,
                          TRUE ~ Time))
```

## all of the data is now merged. We need to filter out bad data points
### Filter 'total' > 0.90 & -0.1 < 
### Find average abundance's and enrichment's
```{r, warning=F}
tidy_data <- full_data_1 %>%
  filter(Total > 0.90) %>%
  group_by(Metabolite, Frag, MW, Time, Bee, Replicate) %>%
  filter(!any(Abundance < -0.01)) %>%
  filter(!any(Abundance > 1.01))

tidy_data$MW <- factor(tidy_data$MW)
tidy_data$MI <- factor(tidy_data$MI)

tech_avg_data <- tidy_data %>%
  select(Metabolite, Frag, MW, MI, Abundance, Enrichment, Time, Bee, Replicate) %>%
  ungroup() %>%
  group_by(Metabolite, Frag, MW, MI, Time, Bee) %>%
  summarise(mean_abundance = mean(Abundance),
            sd_abundance = sd(Abundance),
            mean_Enrichment = mean(Enrichment),
            sd_Enrichment = sd(Enrichment))
  
avg_data <- tech_avg_data %>%
  ungroup() %>%
  group_by(Metabolite, Frag, MW, MI, Time) %>%
  summarise(Mean_abundance = mean(mean_abundance),
            Sd_abundance = sd(mean_abundance),
            Mean_Enrichment = mean(mean_Enrichment),
            Sd_Enrichment = sd(mean_Enrichment)) %>%
  mutate(across(where(is.numeric), round, 3),
         Sd_abundance = case_when(Sd_abundance < 0.005 ~ 0.005,
                                  TRUE ~ Sd_abundance))
# Remove Asp [234] 390; Cit [12345] 431; Glu [2345] 330; Hpro 416
avg_Enrichments <- avg_data %>%
  filter(MI == 0 & Metabolite != "Pyrdc" & Metabolite != "Urate" & Metabolite != "Hpro" & 
           Metabolite != "Gln" &Metabolite != "Fum") %>%
  filter(Metabolite != "Nle") %>%
  filter(MW %in% c("260","418","459","287","431","432","246","315","416",
                   "274","320","336","286","330","390","289","466","567"))%>%
  filter(!((Metabolite == "Asp" & MW == "390") |(Metabolite == "Cit" & MW == "431") | 
           (Metabolite == "Glu" & MW == "330")))%>%
  mutate(Class = case_when(
           Metabolite == "Lac" | Metabolite == "Suc" |  Metabolite == "Fum" | Metabolite == "Glyc" |
             Metabolite == "HMG" | Metabolite == "Mal" | Metabolite == "Cit"  ~ "Carboxylate",
           Metabolite == "Leu" | Metabolite == "Ile" |  Metabolite == "Met" | Metabolite == "Phe" |
             Metabolite == "Hpro" | Metabolite == "Tyr" ~ "EAA",
           Metabolite == "Ala" | Metabolite == "Gly" |  Metabolite == "Pro" | Metabolite == "Ser" |
            Metabolite == "Asp" | Metabolite == "Glu" | Metabolite == "Gln" ~ "NEAA",
           TRUE ~ "other"))
```
## Figure 3D
```{r, warning=FALSE}

ggplot(avg_Enrichments, aes(x = Time, y = Mean_Enrichment))+
  stat_summary(aes(Time, Mean_Enrichment, fill=Class), geom = "ribbon", fun.data = mean_cl_boot, alpha=0.15)+
  guides(fill = "none")+
  #stat_summary(aes(Time, Mean_Enrichment, colour=Class), geom = "line", fun = mean,)+
  geom_point(aes(color = Class),size = 1.5,stroke =0, na.rm=T) +
  geom_line(aes(group=Metabolite, color = Class),size = 0.15,stroke =0, na.rm=F) +
  geom_hline(yintercept=1.08, col="black", linetype = "dashed") +
  theme(axis.title.x = element_text(vjust = -0.6, size=7),
    panel.background = element_rect(fill = "white", colour = "grey50") ,legend.position = c(.7, .75),
        axis.text.x = element_text(color = "black", size=7), 
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        axis.text.y = element_text(color = "black", size=7),
        axis.title.y =element_text(color = "black", size=7),
        legend.text = element_text(color = "black", size=7), 
        legend.title = element_text(color = "black", size=7),
         aspect.ratio=1)+
    scale_colour_manual(labels = c("Carboxylates", "EAA", "NEAA"),  
                        values = c("#6388b4", "#ffae34", "#55ad89"))+ 
  scale_fill_manual(labels = c("Carboxylates", "EAA", "NEAA"),  
                        values = c("#6388b4", "#ffae34", "#55ad89"))+ 
  #old colors:"blue3","darkorange3", "springgreen3"
  scale_y_continuous(limits=c(-2,100), breaks=seq(0, 100, 20))  +
  ylab(expression(""^13*"C"~"Enrichment"~" (%)"))+
    theme(legend.key.size = unit(0.4, 'cm'))+
  scale_x_continuous(name = "Time (h)", limits=c(0,50), breaks=seq(0, 50, 8))+
  annotate("text", x=10, y=6, size=2, label= "Natural Abundance")
#ggsave("Fig3d_revised2.pdf",device=cairo_pdf, width=90, height= 60, units = "mm", dpi = 600)

#write_xlsx(avg_Enrichments, "Dataset_S2.xlsx")

```

## Plot selected mass isotopologue (MI) abundances
```{r, warning=FALSE}

ggplot(subset(avg_data, Metabolite %in% "Ser" & MW %in% c("302", "362", "390")), aes(x = MI, y = Mean_abundance, 
              ymin=Mean_abundance, ymax = Mean_abundance + Sd_abundance, fill = MI))+
  geom_col(color="black") +
  geom_errorbar( width=0.2, color="black") +
  scale_fill_brewer(palette = "Reds") +
  theme(panel.background = element_rect(fill = "white", colour = "grey50"), legend.position = "bottom", 
        axis.text.x = element_text(color = "black", size=12), axis.text.y = element_text(color = "black", size=10),
        axis.title.x =element_text(color = "black", size=16), axis.title.y =element_text(color = "black", size=16),
        legend.text = element_text(color = "black", size=14), legend.title = element_text(color = "black", size=14),
        plot.title = element_text(hjust = 0.5, size=18),
        strip.text.x = element_text(size = 10, color = "black"), 
        strip.text.y = element_text(size = 10, color = "black"),
        strip.background = element_rect(color="grey50", fill="white", size=0.5, linetype="solid"))+
  scale_y_continuous(name = "Abundance",limits=c(0,1), breaks=seq(0, 1, 0.1))  +
  scale_x_discrete(name = "Mass Isotopologue (n+i)")+
  ggtitle("Serine fragments") +
  facet_grid(vars(Frag), vars(Time))
#ggsave("Ser_MIs.pdf")

ggplot(subset(avg_data, Metabolite %in% c("Mal","Suc") & MW %in% c("287","289")), aes(x = MI, y = Mean_abundance, 
              ymin=Mean_abundance, ymax = Mean_abundance + Sd_abundance, fill = MI))+
  geom_col(stat="identity", color="black") +
  geom_errorbar( width=0.2, color="black") +
  scale_fill_brewer(palette = "Reds") +
  theme(panel.background = element_rect(fill = "white", colour = "grey50"), legend.position = "bottom", 
        axis.text.x = element_text(color = "black", size=12), axis.text.y = element_text(color = "black", size=10),
        axis.title.x =element_text(color = "black", size=16), axis.title.y =element_text(color = "black", size=16),
        legend.text = element_text(color = "black", size=14), legend.title = element_text(color = "black", size=14),
        plot.title = element_text(hjust = 0.5, size=18),
        strip.text.x = element_text(size = 10, color = "black"), 
        strip.text.y = element_text(size = 10, color = "black"),
        strip.background = element_rect(color="grey50", fill="white", size=0.5, linetype="solid"))+
  scale_y_continuous(name = "Abundance",limits=c(0,1), breaks=seq(0, 1, 0.1))  +
  scale_x_discrete(name = "Mass Isotopologue (n+i)")+
  ggtitle("Malate | Succinate") +
  facet_grid(vars(Metabolite), vars(Time))

ggplot(subset(avg_data, Metabolite %in% c("Cit","HMG") & MW %in% c("459","315")), aes(x = MI, y = Mean_abundance, 
              ymin=Mean_abundance, ymax = Mean_abundance + Sd_abundance, fill = MI))+
  geom_col(stat="identity", color="black") +
  geom_errorbar( width=0.2, color="black") +
  scale_fill_brewer(palette = "Reds") +
  theme(panel.background = element_rect(fill = "white", colour = "grey50"), legend.position = "bottom", 
        axis.text.x = element_text(color = "black", size=7), 
        axis.text.y = element_text(color = "black", size=7),
        axis.title.x =element_text(color = "black", size=7), 
        axis.title.y =element_text(color = "black", size=7),
        legend.text = element_text(color = "black", size=7), 
        legend.title = element_text(color = "black", size=7),
        plot.title = element_text(hjust = 0.5, size=7),
        strip.text.x = element_text(size = 7, color = "black"), 
        strip.text.y = element_text(size = 7, color = "black"),
        strip.background = element_rect(color="grey50", fill="white", size=0.5, linetype="solid"),
         aspect.ratio=1)+
  scale_y_continuous(name = "Abundance",limits=c(0,1), breaks=seq(0, 1, 0.1))  +
  scale_x_discrete(name = "Mass Isotopologue (n+i)")+
  ggtitle("Citrate | 3Hmg") +
  facet_grid(vars(Metabolite), vars(Time))

```


