---
title: "Bee Gut CFU Counts"
author: "Andrew Quinn, University of Lausanne"
date: "1/11/2022"
output: html_document
keep_tex: yes
always_allow_html: true
---

# Generate plots for folony forming units (CFU's) plated from bee guts, as well as weight calculations. 
## Figure S1A, Figure S4, Figure 5, 



## Environment setup
### Install and load libraries (run code if missing libraries)
#### These are the required package to reproduce this analysis.

```{r, echo = T, eval=T, message=F, warning=F, results=F}


#required_packages = c("readxl",  "tidyverse",    "RColorBrewer" , "ggbeeswarm","ggpubr", "reshape2" , "gridExtra", "stringr","EnvStats", "multcompView", "stats", "ggh4x")
#need_install = required_packages[!(required_packages) %in% installed.packages()]

#if (length(need_install) > 0){
#  install.packages((need_install))
#}

#lapply(required_packages, require, character.only=T)

#dir_script = dirname(rstudioapi::getSourceEditorContext()$path)

#setwd(dir_script)

```


# Plot CFU counts across diet and colonizations groups
```{r, warning=FALSE, message=FALSE}
library(readxl)  
library(tidyverse)
library(RColorBrewer) 
library(ggbeeswarm)
library(ggpubr) 
library(reshape2)  
library(gridExtra) 
library(stringr)
library(EnvStats)
library(multcompView) 
library(stats)
library(cowplot)
library(ggfortify)
library(lme4)
library(nlme)

CFUs <- data.frame(read.csv("All_CFU_Count.csv")) 
CFUs$Colonization <- factor(CFUs$Colonization, levels=c("MF", "Sn", "Gi", "Sn_Gi"))
CFUs$Diet <- factor(CFUs$Diet, levels=c("SW", "P"))
CFUs$Condition <- factor(CFUs$Condition, levels=c("MF_SW", "MF_P", "Sn_SW", "Sn_P","Gi_SW", "Gi_P","Sn_Gi_SW", "Sn_Gi_P"))
CFUs$Experiment <- factor(CFUs$Experiment)

cfu_anova <- aov(log10(CFU)~ Condition, data=CFUs, na.action = na.omit)
tHSD_cfu <- TukeyHSD(cfu_anova, conf.level = 0.95)
#plot(tHSD_cfu , las=1 , col="brown")


generate_label_df <- function(tHSD_cfu, variable){
  
  # Extract labels and factor levels from Tukey post-hoc 
  Tukey.levels <- tHSD_cfu[[variable]][,4]
  Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])
  
  #I need to put the labels in the same order as in the boxplot :
  Tukey.labels$Cage=rownames(Tukey.labels)
  Tukey.labels=Tukey.labels[order(Tukey.labels$Cage) , ]
  return(Tukey.labels)
}
LABELS=generate_label_df(tHSD_cfu , "Condition")

tHSD_label <- LABELS %>%
  mutate(yhsd = 1e+08)

compare_means(CFU ~ Diet, group.by = "Colonization", data = CFUs)

diet_comp <- list( c("Sn_SW", "Sn_P"),c("Gi_SW", "Gi_P"),c("Sn_Gi_SW", "Sn_Gi_P"))

ggplot(data = CFUs, mapping = aes(x = Condition, y = CFU)) +
  geom_quasirandom(fill="grey70",color = "black", height = 0, width = 0.25, shape=21, size = 1.5, stroke=0.25) +
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                  geom = "crossbar",width = 0.5, color="salmon") +
  theme(axis.title.x = element_text(vjust = -0.6, size=7),
    panel.background = element_rect(fill = "white", colour = "grey50") ,   legend.position = "none",
        axis.text.x = element_text(color = "black", size=7), 
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        axis.text.y = element_text(color = "black", size=7),
        axis.title.y =element_text(color = "black", size=7),
        legend.text = element_text(color = "black", size=7), 
        legend.title = element_text(color = "black", size=7),
        plot.title = element_text(hjust = 0.5, size=7),
         aspect.ratio=1)+
  scale_y_log10( name = "CFU's / Gut", breaks = scales::trans_breaks("log10", function(x) 10^x),
                 labels = scales::trans_format("log10", scales::math_format(10^.x)),
                 limits = c(1e+01, 5e+9)) +
  scale_x_discrete(labels=c('MF','MF (P)', 'Sn', 'Sn (P)', 'Gi', 'Gi (P)','Sn+Gi', 'Sn+Gi (P)')) +
  annotation_logticks(sides = "l") +
  geom_hline(yintercept=70, col="grey50", linetype = "dashed") +
  stat_compare_means(comparisons = diet_comp, label =  "p.signif", label.y=c(8,8.5,9)) +
  stat_n_text(na.rm = TRUE, size=2, y.pos = 1)+
  geom_text(aes(x = 3, y = 120), color="black", label = 'LOD', size=2)
 # ggsave("Figure_S1A.pdf" ,device=cairo_pdf, width=90, height= 120, units = "mm", dpi = 600)

  
 (ggplot(subset(CFUs, Colonization %in% "Sn"), mapping = aes(x = Condition, y = CFU)) +
  geom_quasirandom(fill="grey70",color = "black", height = 0, width = 0.25, shape=21, size = 1.5, stroke=0.25) +
     stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                  geom = "crossbar",width = 0.5, color="salmon") +
     stat_summary(fun.y = median, geom = "line", color="grey20",width = 1, aes(group = 1))+
  theme(axis.title.x = element_text(vjust = -0.6, size=8),
    panel.background = element_rect(fill = "white", colour = "grey50") ,   legend.position = "none",
        axis.text.x = element_text(color = "black", size=8), 
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        axis.text.y = element_text(color = "black", size=8),
        axis.title.y =element_text(color = "black", size=8),
        legend.text = element_text(color = "black", size=6), 
        legend.title = element_text(color = "black", size=6),
        plot.title = element_text(hjust = 0.5, size=8))+
  scale_y_log10( name = "CFU's / Gut", breaks = scales::trans_breaks("log10", function(x) 10^x),
                 labels = scales::trans_format("log10", scales::math_format(10^.x)),
                 limits = c(1e+01, 5e+9)) +
  scale_x_discrete(labels=c('Sn', 'Sn (P)')) +
  annotation_logticks(sides = "l") +
  geom_hline(yintercept=70, col="grey50", linetype = "dashed") +
  stat_n_text(na.rm = TRUE, size=2, y.pos = 1)+
  geom_text(aes(x = 1, y = 120), color="black", label = 'LOD', size=2)+
  facet_wrap(~Experiment, nrow=1))
  
  CFU_summary <-CFUs %>%
  group_by(Condition) %>%
  summarise(avg = mean(CFU),
            med = median(CFU),
            stdev = sd(CFU))
```
## Bee weights
```{r, warning=FALSE, message=FALSE}
Weights <- CFUs %>%
  select(-CFU) %>%
  mutate(Bee_weight = 1000*Bee_weight,  # convert g to mg
         Gut_weight = 1000*Gut_weight,
         Body_weight=Bee_weight-Gut_weight)

Weights$Condition <- as.character(Weights$Condition)
Weights$Condition <- factor(Weights$Condition, levels=c("MF_SW", "MF_P", "Sn_SW", "Sn_P","Gi_SW", "Gi_P","Sn_Gi_SW", "Sn_Gi_P"))


p1<- ggplot(data = Weights, mapping = aes(x = Condition, y = Bee_weight)) +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(fill="grey70",color = "black", height = 0, width = 0.15, shape=21, size = 1, stroke=0) +
  theme(text=element_text(size=7,  family="Arial"),
        panel.background = element_rect(fill = "white", colour = "grey50"),
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        axis.text.x = element_text(color = "black", size=5), 
        axis.text.y = element_text(color = "black", size=6),
        axis.title.x =element_text(color = "black", size=7), 
       axis.title.y =element_text(color = "black", size=7),
         aspect.ratio=1)+
  scale_y_continuous( name = "Bee weight (mg)",limits=c(0,200),breaks=seq(0,200,25))+
  stat_n_text(na.rm = TRUE, size=1.5, y.pos = 1) 

p2<- ggplot(data = Weights, mapping = aes(x = Condition, y = Gut_weight)) +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(fill="grey70",color = "black", height = 0, width = 0.15, shape=21, size = 1, stroke=0) +
  theme(text=element_text(size=7,  family="Arial"),
        panel.background = element_rect(fill = "white", colour = "grey50"),
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        axis.text.x = element_text(color = "black", size=5), 
        axis.text.y = element_text(color = "black", size=6),
        axis.title.x =element_text(color = "black", size=7), 
       axis.title.y =element_text(color = "black", size=7),
         aspect.ratio=1)+
  scale_y_continuous( name = "Gut weight (mg)",limits=c(0,125),breaks=seq(0,125,25))+
  stat_n_text(na.rm = TRUE, size=1.5, y.pos = 1) 

p3<- ggplot(data = Weights, mapping = aes(x = Condition, y = Body_weight)) +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(fill="grey70",color = "black", height = 0, width = 0.15, shape=21, size = 1, stroke=0) +
  theme(text=element_text(size=7,  family="Arial"),
        panel.background = element_rect(fill = "white", colour = "grey50"),
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        axis.text.x = element_text(color = "black", size=5), 
        axis.text.y = element_text(color = "black", size=6),
        axis.title.x =element_text(color = "black", size=7), 
       axis.title.y =element_text(color = "black", size=7),
         aspect.ratio=1)+
  scale_y_continuous( name = "Body weight (mg)",limits=c(0,150),breaks=seq(0,150,25))+
  stat_n_text(na.rm = TRUE, size=1.5, y.pos = 1) 

plot_grid(p1, p2, p3, nrow=1)
# ggsave("FigSx_Weights_v2.pdf" ,device=cairo_pdf, width=210, height= 90, units = "mm", dpi = 600)

Weight_summary_strains <-Weights %>%
  group_by(Condition) %>%
  summarise(across(c(Bee_weight,Gut_weight,Body_weight), 
                   .f = list(mean = mean, sd = sd), na.rm = TRUE))

Diet_weight_dif <- Weights %>%
  group_by(Diet)%>%
  summarise(across(c(Bee_weight,Gut_weight,Body_weight), 
                   .f = list(mean = mean, sd = sd), na.rm = TRUE))
```
## Test for diet and colonization effects on weights
```{r, message=FALSE, warning=FALSE}

weight_lmm <- Weights %>%
  select(-Condition) %>%
  mutate(Sn=case_when(str_detect(Colonization,"Sn")~1,
                      TRUE ~ 0),
         Gi=case_when(str_detect(Colonization,"Gi")~1,
                      TRUE ~ 0))%>%
  ungroup() %>%
  select(-Colonization)%>%
  mutate(Diet =as.factor(Diet),
         Sn=as.factor(Sn),
         Gi=as.factor(Gi),
         Experiment=as.factor(Experiment))%>%
  drop_na(Body_weight)

bee_lme_v0 = lme(Bee_weight ~ Diet + Sn*Gi, random = ~1|Experiment, data = weight_lmm, method = "ML")
bee_lme_v1 <- update(bee_lme_v0, weights = varIdent(form = ~ 1 | Experiment))
anova(bee_lme_v0,bee_lme_v1)
summary(bee_lme_v1)
anova(bee_lme_v1)

gut_lme_v0 = lme(Gut_weight ~ Diet + Sn*Gi, random = ~1|Experiment, data = weight_lmm, method = "ML")
gut_lme_v1 <- update(gut_lme_v0, weights = varIdent(form = ~ 1 | Experiment))
anova(gut_lme_v0,gut_lme_v1)
summary(gut_lme_v1)
anova(gut_lme_v1)

body_lme_v0 = lme(Body_weight ~ Diet + Sn*Gi, random = ~1|Experiment, data = weight_lmm, method = "ML")
body_lme_v1 <- update(body_lme_v0, weights = varIdent(form = ~ 1 | Experiment))
anova(body_lme_v0,body_lme_v1)
summary(body_lme_v1)
anova(body_lme_v1)
intervals(body_lme_v1, level = 0.95,which = "fixed")

```

## CFU counts for Figure 5
```{r, warning=FALSE, message=FALSE}

CFUs_strain <- data.frame(read.csv("CFU_Count_AM.csv"))

CFUs_strain$Colony_count <- as.numeric(CFUs_strain$Colony_count)
CFUs_strain$Dilution <- as.numeric(CFUs_strain$Dilution)

CFUs_strain <- CFUs_strain%>%
  mutate(cfu_count = Colony_count * 700 / 10 * 10 ^ Dilution,
         cfu_count = case_when(cfu_count == 70 ~50,
                               TRUE ~ cfu_count)) %>%
  drop_na(cfu_count) 

Success <- CFUs_strain %>%
  group_by(Species, Colonization) %>%
  summarise(n_bees = n(),
            n_col = sum(cfu_count > 70),
            colonized = n_col / n_bees) %>%
  mutate(colonized = scales::percent(colonized, accuracy = 1),
         ypos = 20)

Success_native <- CFUs_strain %>%
  filter(Native != "None") %>%
  group_by(Species, Native) %>%
  summarise(n_bees = n(),
            n_col = sum(cfu_count > 70),
            colonized = n_col / n_bees) %>%
  mutate(colonized = scales::percent(colonized, accuracy = 1),
         n_mf = n_bees - n_col)  

Success_apis <- Success_native %>%
  filter(Species == "Apis") %>%
  select(Native, n_col,n_mf) %>%
  column_to_rownames(var = "Native") %>%
  select(n_col,n_mf)

fisher.test(Success_apis)

cfu_anova <- aov(log10(cfu_count)~ Colonization, data=CFUs_strain, na.action = na.omit)
tHSD_cfu <- TukeyHSD(cfu_anova, conf.level = 0.95)

generate_label_df <- function(tHSD_cfu, variable){
  
  # Extract labels and factor levels from Tukey post-hoc 
  Tukey.levels <- tHSD_cfu[[variable]][,4]
  Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])
  
  #I need to put the labels in the same order as in the boxplot :
  Tukey.labels$Colonization=rownames(Tukey.labels)
  Tukey.labels=Tukey.labels[order(Tukey.labels$Colonization) , ]
  return(Tukey.labels)
}

# Apply the function on my dataset
LABELS=generate_label_df(tHSD_cfu, "Colonization")

CFUs_strain$Colonization <- as.character(CFUs_strain$Colonization)
CFUs_strain$Colonization <- factor(CFUs_strain$Colonization, levels=c('MF',	'wkb2', 'ESL0304',	'ESL0323',	'ESL0251','ESL0897'))

tHSD_label <- LABELS %>%
  mutate(yhsd = 1e+08)

ggplot(data = CFUs_strain, mapping = aes(x = Colonization, y = cfu_count)) +
  geom_hline(yintercept=70, col="grey50", linetype = "dashed", size = 0.75) +
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                  geom = "crossbar",width = 0.5, color="salmon") +
  geom_quasirandom(fill="grey70",color = "black", height = 0, width = 0.15, shape=21, size = 2, stroke=0.25) +
 theme(text=element_text(size=7,  family="Arial"),
        panel.background = element_rect(fill = "white", colour = "grey50"),
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        axis.text.x = element_text(color = "black", size=7), 
        axis.text.y = element_text(color = "black", size=7),
        axis.title.x =element_text(color = "black", size=7), 
       axis.title.y =element_text(color = "black", size=7),
         aspect.ratio=1)+
  scale_y_log10( name = "CFU's / Gut", breaks = scales::trans_breaks("log10", function(x) 10^x),
        labels = scales::trans_format("log10", scales::math_format(10^.x)),limits = c(1e+01, 1e+08)) +
  annotation_logticks(sides = "l") +
  scale_x_discrete(breaks=c('MF',	'wkb2', 'ESL0304',	'ESL0323',	'ESL0251',	'ESL0893',	'ESL0897'), 
                   labels=c('MF',	'wkB2', 'ESL0304',	'ESL0323',	'ESL0251',	'ESL0893',	'ESL0897')) +
  geom_vline(xintercept=4.5, col="grey50", linetype = "dotted") +
  geom_text(data = tHSD_label, aes(x = Colonization,  y = yhsd, label = Letters), size=5) +
  stat_n_text(na.rm = TRUE, size=2, y.pos = 1) +
  geom_text(data = Success, aes(x = Colonization,  y = ypos, label = colonized), size=2) +
  geom_text(aes(x = 3, y = 45), color="black", label = 'Colonization Success', size=2) +
  geom_text(aes(x = 3, y = 120), color="black", label = 'LOD', size=2)
  #ggsave("Figure_5a.pdf", device=cairo_pdf, width=90, height= 120, units = "mm", dpi = 600)

CFU_summary_strains <-CFUs_strain %>%
  group_by(Species, Colonization) %>%
  summarise(avg = mean(cfu_count),
            med = median(cfu_count),
            stdev = sd(cfu_count))  
  
```

