---
title: "NanoSIMS_analysis"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(fig.width=120, fig.height=80) 
```

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

```{r wrap-hook, eval=T}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```

#### These are the required package to reproduce this analysis.

```{r, echo = T, eval=T, message=F, warning=F, results=F}

### env setup

required_packages = c("dplyr", "data.table", "ggplot2", "readxl", "ggbeeswarm", "ggpubr", "EnvStats", "tidyverse", "patchwork", "ggforce", "wesanderson", "RcolorBrewer", "extrafont", "ggthemes" ,"egg", "officer", "rvg","ggh4x",
"grid")
need_install = required_packages[!(required_packages) %in% installed.packages()]

if (length(need_install) > 0){
  install.packages((need_install))
}

# Load packages
lapply(required_packages, require, character.only=T)

dir_script = dirname(rstudioapi::getSourceEditorContext()$path)

setwd(dir_script)
#font_import()
#loadfonts(device="win")

```
  
  

#### **Tidying up the data and adding metadata to the dataset**


##### 1. Calculate the atom % enrichment as well as the delta value for both species measured

##### 2. Produce quality check plots 

##### 3. Produce plot for supplementary and main text figures

```{r, echo = T, eval=T, warning=FALSE, fig.asp = 0.8, fig.width = 7}

## Load the data

dat= fread("NanoSIMS_data.csv")

## Needed only if interested in the 13C12C ratio
#dat[, ratio_13c12c:=`13C 12C /12C2`/mean(`13C 12C /12C2`[Label=="12C"]), ] ## Calculate mean of the ratio 1 

#dat[, delta_13c12c:=(`13C 12C /12C2`/mean(`13C 12C /12C2`[Label=="12C"])-1)*1000, ] ## Calculate delta permil


## Calculate the ratio by dividing the ratio of labeled and labeled samples (12C)
dat[, ratio_14N13c :=dat$`14N 13C /14N 12C` /dat$`14N 13C /14N 12C`[Label=="12C"], ] 

dat$ratio_14N13c

## Calculate delta permil by removing 1 to the ratio and multiplying by 1000

dat[, delta_14N13c:=(`14N 13C /14N 12C`/mean(`14N 13C /14N 12C`[Label=="12C"], na.rm=T)-1)*1000, ] 

dat$delta_14N13c

## calculate the percentage of 13C enrichment in 1 sample
dat[, atpercent_14N13c:=100*(`14N 13C `)/(`14N 13C `+`14N 12C `) ]

## Explore the data 
min(dat$atpercent_14N13c, na.rm=T)

max(dat$atpercent_14N13c, na.rm=T)

head(dat)


```
 
#### Quick check at the correlation between the two measured molecular weights
```{r, echo = T, eval=T, warning=FALSE, fig.asp = 0.8, fig.width = 7, message=FALSE}

## Correlation between the ratio of 14N13C/14N12C and 13C12C/12C2
(ggplot(dat, aes(x=dat$`14N 13C /14N 12C`, dat$`13C 12C /12C2`), colour=dat$Sample)+
  geom_point(mapping = aes(colour=dat$Sample))+
  geom_smooth(method="lm", size=0.5, colour="black")+
  theme_light()+
  scale_y_continuous(breaks = seq(0, 0.15, by = 0.01))+
  scale_x_continuous(breaks = seq(0, 0.1, by = 0.01))+
  ylab("13C12C/12C2 ratio")+
  scale_colour_brewer(palette="Set1"))+
  labs(colour='Sample') +
  ggtitle("Ratio correlation")+
  xlab("14N13C/14N12C ratio")+
  stat_cor()+
  theme(text=element_text(family="Arial", face="bold", size=12))

ggsave("Ratios_correlation.pdf", device=cairo_pdf)

## Correlation between the counts of 14N12C and 12C2

(ggplot(dat, aes(x=dat$`12C2 `, dat$`14N 12C `), colour=dat$Sample)+
  geom_point(mapping = aes(colour=dat$Sample))+
  geom_smooth(method="lm", size=0.5, colour="black")+
  theme_light()+
  scale_y_continuous(breaks = seq(0, 0.15, by = 0.01))+
  scale_x_continuous(breaks = seq(0, 0.1, by = 0.01))+
  ylab("14N12C counts")+
  scale_colour_brewer(palette="Set1"))+
  labs(colour='Sample') +
  ggtitle("Species correlation")+
  xlab("12C2 counts")+
  stat_cor()+
  theme(text=element_text(family="Arial", face="bold", size=12))

```


#### Quick check of different metrics to assess acquisition batch and experiment effects
```{r, echo = T, eval=T, warning=FALSE, fig.asp = 0.8, fig.width = 7}

## Evaluation of error per batch
(ggplot(dat, aes(x=as.factor(dat$`Date acquired`), dat$`ER-14N 13C /14N 12C`, colour=dat$Tissue))+
  geom_quasirandom(dodge.width = 0.5)+
  theme_light()+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  labs(colour='Sample') +
  scale_y_continuous(breaks = seq(0, 0.0016, by = 0.0001))+
  ylab("14N13C/14N12C error")+
  xlab("Batch")+
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial", face="bold", size=12))
)

## Evaluation of 14N12C counts per Acquisition date

(ggplot(dat, aes(x=as.factor(dat$`Date acquired`), dat$`14N 12C `))+
  geom_boxplot()+
  geom_quasirandom(dodge.width = 0.5)+
  theme_light()+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  labs(colour='Sample') +
  ylab("14N12C counts")+
  xlab("Date acquired")+
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial", face="bold", size=12))
)

## Evaluation of the 14N13C counts per Acquisition date
(ggplot(dat, aes(x=as.factor(dat$`Date acquired`), dat$`14N 13C `))+
  geom_boxplot(aes(group=`Date acquired`))+
  geom_quasirandom(dodge.width = 0.5)+
  theme_light()+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  labs(colour='Sample') +
  ylab("14N13C")+
  xlab("Date acquired")+
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial", face="bold", size=12))
)


## Evaluation of 12C2 counts per Acquisition date
(ggplot(dat, aes(x=as.factor(dat$`Date acquired`), dat$`12C2 `))+
  geom_boxplot(aes(group=`Date acquired`))+
  geom_quasirandom(dodge.width = 0.5)+
  theme_light()+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  labs(colour='Sample') +
  ylab("12C2 counts")+
  xlab("Date acquired")+
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial", face="bold", size=12))
)

## Evaluation of the 14N12C counts per sample and coloured by tissue type
(ggplot(dat, aes(x=dat$Sample, dat$`14N 12C `, colour=Tissue))+
  geom_quasirandom(dodge.width = 0.5)+
  theme_light()+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  labs(colour='Sample') +
  ylab("14N12C counts")+
  xlab("Sample")+
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial", face="bold", size=12))+
  facet_wrap(~Label)
)

## sum of 14N12C and 14N13C per image file
(ggplot(dat, aes(x=as.character(dat$IMAGEFILE), y=dat$`14N 12C `+dat$`14N 13C `))+
  geom_quasirandom(dodge.width = 0.5)+
  theme_light()+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  labs(colour='Sample') +
  ylab("14N12C+14N13C counts")+
  xlab("Image")+  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial", face="bold", size=12))
)

## sum of 14N12C and 14N13C per acqusition date
(ggplot(dat, aes(x=as.character(dat$`Date acquired`), y=dat$`14N 12C `+dat$`14N 13C `))+
  geom_boxplot()+
  geom_quasirandom(dodge.width = 0.5)+
  theme_light()+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  labs(colour='Sample') +
  ylab("14N12C+14N13C counts")+
  xlab("Date acquried")+  
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial", face="bold", size=12))
)

## sum of 14N12C and 14N13C per Experiment
(ggplot(dat, aes(x=as.character(dat$Experiment), y=dat$`14N 12C `+dat$`14N 13C `))+
  geom_boxplot()+
  geom_quasirandom(dodge.width = 0.5)+
  theme_light()+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  labs(colour='Sample') +
  ylab("14N12C+14N13C counts")+
  xlab("Experiment")+  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial", face="bold", size=12))+
  stat_compare_means(method="kruskal")
)


```
#### Biggest batch effect is between the two experiments. 

#### ROI sizes for host and for bacteria for the 12C and 13C ROIs
```{r, echo = T, eval=T, warning=FALSE, fig.asp = 0.8, fig.width = 7}

## Subset the 12C and 13C samples in distinct tables
dat_12C = dat[Label == "12C"]
dat_13C = dat[Label == "13C"]

## Plot the sizes of all ROIS of the 12C samples
(ggplot(dat_12C, aes(x=dat_12C$Tissue, dat_12C$ROIAREA))+
  geom_boxplot()+
  ylab("ROI area")+
  xlab("Tissue type")+
  geom_quasirandom(aes(colour=Experiment))+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  stat_n_text()+
  theme_light()+
  facet_zoom(ylim = c(0, 2.5))+
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial",  size=14))
)
ggsave("ROIarea_12C.pdf", device=cairo_pdf)

## Plot the sizes of all ROIS of the 13C samples
(ggplot(dat_13C, aes(x=dat_13C$Tissue, dat_13C$ROIAREA))+
  geom_boxplot()+
  ylab("ROI area")+
  xlab("Tissue type")+
  geom_quasirandom(aes(colour=as.factor(Experiment)))+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  stat_n_text()+
  theme_light()+
  facet_zoom(ylim = c(0, 2.5))+
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial",  size=14))
)

ggsave("ROIarea_13C.pdf", device=cairo_pdf)

```



#### Plots of  all ROI ratios of the controls per tissue type. Quick review of at% enrichment in the different samples

```{r, echo = T, eval=T, warning=FALSE, fig.asp = 0.8, fig.width = 7}

## Subset 12C values only to determine the natural 13C abundance in our samples
dat_12C = dat[Label=="12C"]

## Calulate the mean for 13C12C/12C ratio in 12C samples
dat_12C[, by=.(Tissue), Average_ratio_C := mean(`13C 12C /12C2`, na.rm=T)]


## Unique values for Tissue and avergae ratio
unique(dat_12C[, c("Tissue", "Average_ratio_C")]) ## for wkb2 the ratio seems slightly lower that for host tissues

## calculate the mean of  14N13C/14N12C ration in 12C samples
dat_12C[, by=.(Tissue), Average_ratio_N := mean(`14N 13C /14N 12C`, na.rm=T)]
head(dat_12C)

## Unique values for Tissue and avergae ratio
unique(dat_12C[, c("Tissue", "Average_ratio_N")]) ## similar ratios across the 3 entities


(ggplot(dat_12C, aes(x=dat_12C$Tissue, dat_12C$`14N 13C /14N 12C`))+
  geom_quasirandom(width = 0.1)+
  theme_light()+
  ylab("14N13C/14N12C ratio")+
  xlab("Tissue type")+
  scale_y_continuous(labels = scales::number_format(accuracy = 0.00001))+
  geom_point(stat="summary", fun.y="mean", type=15, colour="red")+
  geom_errorbar(width =0.5, stat="summary", 
    fun.ymin=function(x) {mean(x)-2*sd(x)}, 
    fun.ymax=function(x) {mean(x)+2*sd(x)})+
  stat_compare_means(method = "kruskal")+
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(family="Arial",  size=14))
)
## More variability in bacterial cells compared to the rest
ggsave("standard_ratio_12C_samples.pdf", device=cairo_pdf)

## Overview of the enrichments in the different tissue types
(ggplot(dat, aes(Sample, dat$atpercent_14N13c))+
  geom_boxplot()+
  ylab("At% 13C")+ 
  labs(colour='Tissue') +
  scale_y_continuous(breaks = seq(0, 9, by = 1))+
  theme_light()+
  ggtitle("13C atomic enrichment")+
  scale_fill_manual(values = wes_palette("BottleRocket2"))+
  xlab("Sample"))+
  theme(axis.text.x=element_text(angle=90, hjust=1))+
  theme(text=element_text(size=14,  family="Arial"))+
  facet_wrap(~Tissue)

ggsave( "13C_atomic_ernichment_14N13C.pdf")

```

#### The mean ratio for 13C12C/12C2 in host is 0.02178112 and for bacteria is 0.02191010

#### The mean ratio for 14N13C/14N12C in host is 0.01126608 and for bacteria is 0.01125358


#### To make the two experiments comparable, we synchronized the incubation times by setting the zero to the moment when 13C glucose was replaced to 12C glucose in the diet. This means that for the july replicate that 24 hours are subtracted from the time colum and the column switch time is created with the new timing. 
```{r, echo = T, eval=T, warning=FALSE, fig.asp = 0.8, fig.width = 7}

## Rescale the time between the 2 experiments
dat[Experiment == "Replicate August", `:=`(Switch_time, as.numeric(Time) - 24), ][Experiment == "Replicate May", `:=`(Switch_time, as.numeric(Time))]

## Subset the 12C and 13C samples in distinct tables
dat_12C = dat[Label == "12C"]
dat_13C = dat[Label == "13C"]

## Check the labelling of the host cells and the epithelium with time after switch from 13C glc to 12C glc
(dat_13C %>% subset(., Tissue %in% c("Host cells", "Epithelium"))%>%
ggplot(., aes(as.numeric(Switch_time), as.numeric(atpercent_14N13c)))+
  geom_beeswarm( aes(colour=Experiment))+
  ylab(expression("At% "~""^13*"C"~"Enrichment"))+
  labs(colour='Experiment') +
  theme_base()+
    stat_cor(method = "pearson")+
  geom_smooth(method = "lm", se=TRUE, aes(group=Tissue))+
  scale_y_continuous(limits = c(0,5), breaks = seq(0, 5, by = 1))+
  scale_colour_manual(values = wes_palette("BottleRocket2"))+
  xlab("Time (h)")+
  geom_hline(yintercept = 1.11, linetype="dashed")+
  scale_x_continuous(limits = c(0,168), breaks= seq(0,168,24))+
  theme(strip.text = element_text(size=20), text=element_text(size=14,  family="Arial")
    ,axis.title.x =element_text(face="bold"),
  axis.title.y =element_text(face="bold") )+
  stat_cor(label.y = 8)+
  facet_wrap(~Tissue)
)
ggsave("cuticle_tissue_timecourse.pdf", device=cairo_pdf)


### Testing the drop in labelling with linear models

lm1 = dat[Tissue=="Host cells"&Label=="13C", .(unique(Switch_time),mean(atpercent_14N13c)), by = "Switch_time"] %>% lm(V1~V2, data = .)

lm2 = dat[Tissue=="Epithelium"&Label=="13C", .(unique(Switch_time),mean(atpercent_14N13c)), by = "Switch_time"] %>% lm(V1~V2, data = .)


summary(lm1)
summary(lm2)
anova(lm1,lm2)
```

```{r, echo = T, eval=T, warning=FALSE, fig.asp = 0.8, fig.width = 7}

## Figure 3C, labbeling timeline
(ggplot(subset(dat, Label %in% c("13C")), aes(x=Switch_time, atpercent_14N13c, colour=Tissue))+
  stat_summary(aes(Switch_time, atpercent_14N13c), geom = "pointrange", fun.y = mean, shape = 16, size = 0.2,alpha =0.5,
  fun.ymin = function(x) mean(x) - sd(x), fun.ymax = function(x) mean(x) + sd(x) ) + 
  stat_summary(aes(Switch_time, atpercent_14N13c, colour=Tissue), geom = "line", size =0.5, fun = mean, alpha =0.5)+
  geom_quasirandom(aes(shape=Experiment),alpha =1, size=0.5, colour="black", position=position_dodge2(width =2))+
  scale_colour_manual(labels = c("Epithelium","Host tissue", "S. alvi" ), values = c("blue3", "darkorange3", "springgreen3"))+
  scale_fill_manual( labels = c("Epithelium","Host tissue", "S. alvi"), values = c("blue3", "darkorange3", "springgreen3"))+
  labs(colour='Tissue') + 
  ylab(expression("At% "~""^13*"C"~"Enrichment"))+
  scale_y_continuous(limits = c(0,7), breaks = seq(0, 7, by = 1))+
  scale_x_continuous(limits = c(0,168), breaks = seq(0, 168, by = 8))+
  xlab("Time (h)")+
  annotate("text", x=18.5, y=1.3, size=2, label= "Natural Abundance")+
  #facet_wrap(~Tissue, nrow = 3)+
  geom_hline(yintercept = 1.11, linetype="dashed")+
  theme(axis.text.x=element_text(angle = 90), legend.key.size = unit(0.4, 'cm'))+
     #theme(plot.margin = unit(c(0.2,0.2,0.2,0.2), "cm"), legend.key.size = unit(0.5,"line"))+
  guides(shape = "none")+
  theme(text=element_text(size=8,  family="Arial"),
        legend.text = element_text(color = "black", size=6), 
        legend.title = element_text(color = "black", size=6),
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        panel.background = element_rect(fill = "white", colour = "grey50"),
        legend.position = c(.7, .75), 
        axis.text.x = element_text(color = "black", size=8), 
        axis.text.y = element_text(color = "black", size=8), 
        axis.title.y=element_text(color = "black", size=8),
        axis.ticks.x = element_line(linetype = "dashed"))+
  force_panelsizes(rows = unit(1.6, "in"),
                   cols = unit(2.4, "in")))
  

ggsave("Enrichment_plot_facets_Fig3C.pdf",device=cairo_pdf, width=3, height= 2, units = "in", dpi = 300)

```

#### Scale calcluation for main text figure 4
``` {r, echo = T, eval=T, warning=FALSE, fig.asp = 0.8, fig.width = 7}

##14N13C ratios
scale_function= function(ratio){
delta= ((ratio/0.01125358)-1)*1000
scale_value= ((delta+1000)/(delta+1000+(1000/0.01125358)))*100
return(scale_value)
}


scale_function= function(ratio){
scale_value= (ratio*1.11)/0.01125358
return(scale_value)
}

original_scales = c(0, 0.0113, 0.0158, 0.0223, 0.0315, 0.0441, 0.0621, 0.0874, 0.1229)
atpercent_scale= scale_function(original_scales)
round(atpercent_scale, 2)

## Average Host cells enrichment per sample
dat[Tissue== "Host cells" & Label=="13C", .(mean(atpercent_14N13c), sd(atpercent_14N13c)), by = "Sample" ]

## Average Host cells enrichment overall time points and samples
dat[Tissue== "Host cells" & Label=="13C", .(mean(atpercent_14N13c), sd(atpercent_14N13c)),  ]

## Average epithelium enrichment overall time points and samples
dat[Label=="13C", averageatpercent := mean(atpercent_14N13c, na.rm=T),  by = c("Sample", "Tissue") ]

  
```


