---
title: "Bee Gut Metabolomics"
author: "Andrew Quinn"
date: "25/01/2021"
output: md_document
always_allow_html: true
---
# Analysis of gut metabolomics dataset for (Figure 1B, 1C, S3)
## Experimental Design: Colonized bees [S. alvi, Gilliamella, S. alvi + Gilliamella, MD] fed [sugar water, sugar water + pollen]
### Aliquots from centrifuged gut supernatants were divided to extract soluble metabolites and short chain fatty acids (See Methods). Derivatized samples were run on an Agilent 8890/5977B series GC-MSD with VF-5MS column. Features were extracted using MassHunter Quantitative analysis software (Agilent) and identified using NIST 17 MS library or analytical standards. 


### Load the required package to reproduce this analysis (Need to remove #'s)
### Skip this chunk if required packages already downloaded

```{r, echo = T, eval=T, message=F, warning=F, results=F}

#required_packages = c("dplyr", "data.table", "ggplot2", "readxl", "ggbeeswarm", "ggpubr", #"EnvStats", "tidyverse", "patchwork", "ggforce", "wesanderson", "RcolorBrewer", "extrafont", #"ggthemes" ,"egg", "officer", "rvg","ggh4x", "lubridate", "stringr", "modelr",
#"grid", "lmm2met", "ggrepel")
#need_install = required_packages[!(required_packages) %in% installed.packages()]

#if (length(need_install) > 0){
#  install.packages((need_install))
#}

#lapply(required_packages, require, character.only=T)

#dir_script = dirname(rstudioapi::getSourceEditorContext()$path)

#setwd(dir_script)
```

## Load data sheets
### Convert variables to factors
```{r, message = FALSE,warning = FALSE}
library(readxl)
library(RColorBrewer)
library(EnvStats)
library(ggpubr)
library(lubridate)
library(stringr)
library(modelr)
library(ggforce)
library(tidyverse)

ISTD <- read_csv("ISTD_all.csv")

dat_smet <- read_csv("Data_all.csv")

dat_smet$Bee_exp <- as.factor(dat_smet$Bee_exp)
dat_smet$Condition <- as.factor(dat_smet$Condition)
dat_smet$Food <- as.factor(dat_smet$Food)
dat_smet$Metabolite <- as.factor(dat_smet$Metabolite)
dat_smet$Date <- mdy_hm(dat_smet$Date)
dat_smet$SCFA_Batch <- as.factor(dat_smet$SCFA_Batch)
dat_smet$Smet_Batch <- as.factor(dat_smet$Smet_Batch)

```


## Normalize Response to internal standards
## Data split between GC-MS analysis of soluble metabolites vs short chain fatty acids
```{r, message = FALSE, warning=FALSE}
dat_smet_norm <- dat_smet %>%
  left_join(ISTD, by = c("Sample","Type","Condition","SCFA_Batch","Smet_Batch"))%>%
  mutate(Norm_resp = case_when(
    str_detect(Metabolite, "Acetate") ~ Response / Ival_norm,
    str_detect(Metabolite, "Butyrate") ~ Response / Ival_norm,
    str_detect(Metabolite, "Formate") ~ Response / Ival_norm,
    str_detect(Metabolite, "Propionate") ~ Response / Ival_norm,
    str_detect(Metabolite, "Isovalerate") ~ Response / Ival_norm,
    str_detect(Metabolite, "Maleate") ~ Response / Ival_norm,
    str_detect(Metabolite, "Malonate") ~ Response / Ival_norm,
    str_detect(Metabolite, "Pentanoate") ~ Response / Ival_norm,
    TRUE ~ Response / Nle_norm),
    Metab_type = case_when(
    str_detect(Metabolite, "Acetate") ~ "SCFA",
    str_detect(Metabolite, "Butyrate") ~ "SCFA",
    str_detect(Metabolite, "Formate") ~ "SCFA",
    str_detect(Metabolite, "Propionate") ~ "SCFA",
    str_detect(Metabolite, "Isovalerate") ~ "SCFA",
    str_detect(Metabolite, "Maleate") ~ "SCFA",
    str_detect(Metabolite, "Malonate") ~ "SCFA",
    str_detect(Metabolite, "Pentanoate") ~ "SCFA",
    TRUE ~ "Smet"),
    Decision_Nle = replace_na(Decision_Nle,"ok"),
    Decision_Ival = replace_na(Decision_Ival,"ok"))

Results_SCFA <- dat_smet_norm %>%
  filter(Type == "Sample" | Type == "PE" ) %>%
  filter(Decision_Ival != "error" & Metab_type == "SCFA") 
Results_Smet <- dat_smet_norm %>%
  filter(Type == "Sample" | Type == "PE" ) %>%
  filter(Decision_Ival != "error" & Metab_type == "Smet") 
Results <- bind_rows(Results_Smet,Results_SCFA)

Results_contaminated <- filter(Results, Bee_exp == "E") ## Remove contaminated batch

# adjust pollen extract values to levels in the gut
Results_NC <- filter(Results, Bee_exp != "E") %>%
  filter(Sample != "QC9_PE10" & Sample != "QC8_PE10")%>%
  mutate(Norm_resp=case_when(Condition == "PE"~4.3*Norm_resp,
                             TRUE ~Norm_resp))
```


## Calculate average normalized responses by treatment group
## Calculate fold changes and p-values vs reference group (MD)
```{r, message = FALSE, warning=FALSE}
results_group <- Results_NC %>%
  group_by(Condition, Metabolite) %>%
  summarise(med = median(Norm_resp),
            sd = sd(Norm_resp),
            n = n())

results_group_wide <- results_group %>%
  pivot_wider(names_from = Condition, values_from = c(med, sd, n))

results_group_wide_Sn_SW <- results_group %>%
  filter(Condition == "Sn_SW" | Condition == "MD_SW") %>%
  pivot_wider(names_from = Condition, values_from = c(med, sd, n)) %>%
  mutate(FC = med_Sn_SW / med_MD_SW,
         Log_FC = log2(FC),
         stdev_fc = FC*sqrt((sd_Sn_SW/med_Sn_SW)^2 + (sd_MD_SW/med_MD_SW)^2))

results_group_wide_Sn_P <- results_group %>%
  filter(Condition == "Sn_P" | Condition == "MD_P") %>%
  pivot_wider(names_from = Condition, values_from = c(med, sd, n)) %>%
  mutate(FC = med_Sn_P / med_MD_P,
         Log_FC = log2(FC),
         stdev_fc = FC*sqrt((sd_Sn_P/med_Sn_P)^2 + (sd_MD_P/med_MD_P)^2))

results_group_wide_Gi_SW <- results_group %>%
  filter(Condition == "Gi_SW" | Condition == "MD_SW") %>%
  pivot_wider(names_from = Condition, values_from = c(med, sd, n)) %>%
  mutate(FC = med_Gi_SW / med_MD_SW,
         Log_FC = log2(FC),
         stdev_fc = FC*sqrt((sd_Gi_SW/med_Gi_SW)^2 + (sd_MD_SW/med_MD_SW)^2))

results_group_wide_Gi_P <- results_group %>%
  filter(Condition == "Gi_P" | Condition == "MD_P") %>%
  pivot_wider(names_from = Condition, values_from = c(med, sd, n)) %>%
  mutate(FC = med_Gi_P / med_MD_P,
         Log_FC = log2(FC),
         stdev_fc = FC*sqrt((sd_Gi_P/med_Gi_P)^2 + (sd_MD_P/med_MD_P)^2))

results_group_wide_Sn_Gi_SW <- results_group %>%
  filter(Condition == "Sn_Gi_SW" | Condition == "MD_SW") %>%
  pivot_wider(names_from = Condition, values_from = c(med, sd, n)) %>%
  mutate(FC = med_Sn_Gi_SW / med_MD_SW,
         Log_FC = log2(FC),
         stdev_fc = FC*sqrt((sd_Sn_Gi_SW/med_Sn_Gi_SW)^2 + (sd_MD_SW/med_MD_SW)^2))

results_group_wide_Sn_Gi_P <- results_group %>%
  filter(Condition == "Sn_Gi_P" | Condition == "MD_P") %>%
  pivot_wider(names_from = Condition, values_from = c(med, sd, n)) %>%
  mutate(FC = med_Sn_Gi_P / med_MD_P,
         Log_FC = log2(FC),
         stdev_fc = FC*sqrt((sd_Sn_Gi_P/med_Sn_Gi_P)^2 + (sd_MD_P/med_MD_P)^2))

results_group_wide_MD_P <- results_group %>%
  filter(Condition == "MD_P" | Condition == "MD_SW") %>%
  pivot_wider(names_from = Condition, values_from = c(med, sd, n)) %>%
  mutate(FC = med_MD_P / med_MD_SW,
         Log_FC = log2(FC),
         stdev_fc = FC*sqrt((sd_MD_P/med_MD_P)^2 + (sd_MD_SW/med_MD_SW)^2))

results_group_wide_PE <- results_group %>%
  filter(Condition == "PE" | Condition == "MD_P") %>%
  pivot_wider(names_from = Condition, values_from = c(med, sd, n)) %>%
  mutate(FC = med_PE / med_MD_P,
         Log_FC = log2(FC),
         stdev_fc = FC*sqrt((sd_PE/med_PE)^2 + (sd_MD_P/med_MD_P)^2))

p_value_sw <- Results_NC %>%
  filter(Food == "SW") 

p_value_sn_sw <- p_value_sw %>%
  filter(Condition == "Sn_SW" | Condition == "MD_SW" ) 
p_value_sn_sw <- compare_means(Norm_resp ~ Condition, data = p_value_sn_sw, 
                group.by = "Metabolite", ref.group = "MD_SW", p.adjust.method = "BH")
p_value_sn_sw <- p_value_sn_sw %>%
    mutate(Log10_p = -log10(p)) %>%
  rename(Condition = group2, Control = group1)

p_value_gi_sw <- p_value_sw %>%
  filter(Condition == "Gi_SW" | Condition == "MD_SW" ) 
p_value_gi_sw <- compare_means(Norm_resp ~ Condition, data = p_value_gi_sw, 
                group.by = "Metabolite", ref.group = "MD_SW", p.adjust.method = "BH")
p_value_gi_sw <- p_value_gi_sw %>%
    mutate(Log10_p = -log10(p)) %>%
  rename(Condition = group2, Control = group1)

p_value_sn_gi_sw <- p_value_sw %>%
  filter(Condition == "Sn_Gi_SW" | Condition == "MD_SW" ) 
p_value_sn_gi_sw <- compare_means(Norm_resp ~ Condition, data = p_value_sn_gi_sw, 
                group.by = "Metabolite", ref.group = "MD_SW", p.adjust.method = "BH")
p_value_sn_gi_sw <- p_value_sn_gi_sw %>%
    mutate(Log10_p = -log10(p)) %>%
  rename(Condition = group2, Control = group1)

p_value_p <- Results_NC %>%
  filter(Food == "P") 

p_value_sn_p <- p_value_p %>%
  filter(Condition == "Sn_P" | Condition == "MD_P" ) 
p_value_sn_p <- compare_means(Norm_resp ~ Condition, data = p_value_sn_p, 
                group.by = "Metabolite", ref.group = "MD_P", p.adjust.method = "BH")
p_value_sn_p <- p_value_sn_p %>%
    mutate(Log10_p = -log10(p)) %>%
  rename(Condition = group2, Control = group1)

p_value_gi_p <- p_value_p %>%
  filter(Condition == "Gi_P" | Condition == "MD_P" ) 
p_value_gi_p <- compare_means(Norm_resp ~ Condition, data = p_value_gi_p, 
                group.by = "Metabolite", ref.group = "MD_P", p.adjust.method = "BH")
p_value_gi_p <- p_value_gi_p %>%
    mutate(Log10_p = -log10(p)) %>%
  rename(Condition = group2, Control = group1)

p_value_sn_gi_p <- p_value_p %>%
  filter(Condition == "Sn_Gi_P" | Condition == "MD_P" ) 
p_value_sn_gi_p <- compare_means(Norm_resp ~ Condition, data = p_value_sn_gi_p, 
                group.by = "Metabolite", ref.group = "MD_P", p.adjust.method = "BH")
p_value_sn_gi_p <- p_value_sn_gi_p %>%
    mutate(Log10_p = -log10(p)) %>%
  rename(Condition = group2, Control = group1)

p_value_md <- Results_NC %>%
  filter(Condition == "MD_SW" | Condition == "MD_P") 
p_value_md <- compare_means(Norm_resp ~ Condition, data = p_value_md, 
                group.by = "Metabolite", ref.group = "MD_SW", p.adjust.method = "BH")
p_value_md <- p_value_md %>%
    mutate(Log10_p = -log10(p)) %>%
  rename(Condition = group2, Control = group1)

p_value_pe <- Results_NC %>%
  filter(Condition == "PE" | Condition == "MD_P") %>%
  filter(Metabolite != "Glutarate" & Metabolite != "Propanoate")
p_value_pe <- compare_means(Norm_resp ~ Condition, data = p_value_pe, 
                group.by = "Metabolite", ref.group = "MD_P", p.adjust.method = "BH")
p_value_pe <- p_value_pe %>%
    mutate(Log10_p = -log10(p)) %>%
  rename(Condition = group2, Control = group1)

results_group_wide_Sn_SW <- merge(results_group_wide_Sn_SW, p_value_sn_sw, by=c("Metabolite")) %>%
  rename(med_Condition = med_Sn_SW, sd_Condition = sd_Sn_SW, n_Condition = n_Sn_SW, 
         med_Control = med_MD_SW, sd_Control = sd_MD_SW, n_Control = n_MD_SW)

results_group_wide_Gi_SW <- merge(results_group_wide_Gi_SW, p_value_gi_sw, by=c("Metabolite")) %>%
  rename(med_Condition = med_Gi_SW, sd_Condition = sd_Gi_SW, n_Condition = n_Gi_SW, 
         med_Control = med_MD_SW, sd_Control = sd_MD_SW, n_Control = n_MD_SW)

results_group_wide_Sn_Gi_SW <- merge(results_group_wide_Sn_Gi_SW, p_value_sn_gi_sw, by=c("Metabolite")) %>%
  rename(med_Condition = med_Sn_Gi_SW, sd_Condition = sd_Sn_Gi_SW, n_Condition = n_Sn_Gi_SW, 
         med_Control = med_MD_SW, sd_Control = sd_MD_SW, n_Control = n_MD_SW)

results_group_wide_Sn_P <- merge(results_group_wide_Sn_P, p_value_sn_p, by=c("Metabolite")) %>%
  rename(med_Condition = med_Sn_P, sd_Condition = sd_Sn_P, n_Condition = n_Sn_P, 
         med_Control = med_MD_P, sd_Control = sd_MD_P, n_Control = n_MD_P)

results_group_wide_Gi_P <- merge(results_group_wide_Gi_P, p_value_gi_p, by=c("Metabolite")) %>%
  rename(med_Condition = med_Gi_P, sd_Condition = sd_Gi_P, n_Condition = n_Gi_P, 
         med_Control = med_MD_P, sd_Control = sd_MD_P, n_Control = n_MD_P)

results_group_wide_Sn_Gi_P <- merge(results_group_wide_Sn_Gi_P, p_value_sn_gi_p, by=c("Metabolite")) %>%
  rename(med_Condition = med_Sn_Gi_P, sd_Condition = sd_Sn_Gi_P, n_Condition = n_Sn_Gi_P, 
         med_Control = med_MD_P, sd_Control = sd_MD_P, n_Control = n_MD_P)

results_group_wide_MD_P <- merge(results_group_wide_MD_P, p_value_md, by=c("Metabolite")) %>%
  rename(med_Condition = med_MD_P, sd_Condition = sd_MD_P, n_Condition = n_MD_P, 
         med_Control = med_MD_SW, sd_Control = sd_MD_SW, n_Control = n_MD_SW)

results_sig <- rbind(results_group_wide_Sn_SW,results_group_wide_Sn_P)
results_sig <- rbind(results_sig,results_group_wide_Gi_SW)
results_sig <- rbind(results_sig,results_group_wide_Gi_P)
results_sig <- rbind(results_sig,results_group_wide_Sn_Gi_SW)
results_sig <- rbind(results_sig,results_group_wide_Sn_Gi_P)
results_sig <- rbind(results_sig,results_group_wide_MD_P)

Substrates <- data.frame(Metabolite = c("3-Hydroxy-3-methylglutarate", "alpha-Ketoglutarate", "Citrate", "Gamma-Aminobutyrate", "Glycerate", "Acetate", "Malate", "Succinate", "Fumarate", "Pyruvate", "Hypoxanthine", "Lactate", "Caffeic_acid"),
                        Substrate= c("Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes"))

results_sig_PE <- merge(results_group_wide_PE, p_value_pe, by=c("Metabolite")) %>%
  rename(med_Condition = med_PE, sd_Condition = sd_PE, n_Condition = n_PE, 
         med_Control = med_MD_P, sd_Control = sd_MD_P, n_Control = n_MD_P)%>%
  left_join(Substrates, by = "Metabolite")%>%
  mutate(Substrate=replace_na(Substrate,"No"))

results_sig$Abundance <- "NO"
results_sig$Abundance[results_sig$Log_FC > 0.6 & results_sig$p < 0.05] <- "UP"
results_sig$Abundance[results_sig$Log_FC < -0.6 & results_sig$p < 0.05] <- "DOWN"
  
results_sig$siglabel <- NA
results_sig$Metabolite <- as.character(results_sig$Metabolite)
results_sig$siglabel[results_sig$p < 0.05] <- results_sig$Metabolite[results_sig$p < 0.05]
  
results_sig$siglabel <- factor(results_sig$siglabel)

results_sig$Condition <- factor(results_sig$Condition, 
                                     levels = c("MD_P", "Sn_SW", "Sn_P", "Gi_SW", "Gi_P", "Sn_Gi_SW", "Sn_Gi_P"))
results_sig$Abundance <- factor(results_sig$Abundance, 
                                     levels = c("UP", "DOWN", "NO"))
Host <- results_sig %>%
  filter(Condition == "Sn_SW" | Condition == "Sn_P")%>%
  filter(Abundance == "DOWN")

Diet <- results_sig %>%
  filter(Condition == "MD_P" | Condition == "Sn_P")
  
Microbe <- results_sig %>%
  filter(Condition %in% c("Sn_Gi_SW","Sn_Gi_P","Gi_SW","Gi_P"))


results_sig_sn <- results_sig %>%
  filter(Condition == "Sn_SW" | Condition == "Sn_P" ) 

results_sig_1 <- results_sig %>%
  filter(Condition != "MD_P") %>%
  mutate(Food = case_when(str_detect(Condition, "_P") ~ "Pollen",
                          TRUE ~ "Sugar water"),
         Colonization = case_when(str_detect(Condition, "Sn_SW") ~ "Sn",
                                  str_detect(Condition, "Sn_P") ~ "Sn",
                                  str_detect(Condition, "Sn_Gi_SW") ~ "Sn+Gi",
                                  str_detect(Condition, "Sn_Gi_P") ~ "Sn+Gi",
                                  TRUE ~ "Gi" ),
         Food = fct_relevel(Food, "Sugar water", "Pollen"),
         Colonization = fct_relevel(Colonization, "Sn", "Sn+Gi", "Gi"))%>%
  ungroup() %>%
    add_count(siglabel, name = "Shared")
  results_sig_1$Shared[results_sig_1$Shared == 1236] <- 0
    results_sig_1 <- results_sig_1 %>%
      mutate(siglabel = case_when(Shared > 2 ~siglabel, TRUE ~ NA_integer_))

Sig_count <- results_sig %>%
  group_by(Condition) %>%
  summarise(Total = n(),
            Higher = sum(Abundance == "UP")/Total,
            Lower = sum(Abundance == "DOWN")/Total,
            Same = sum(Abundance == "NO")/Total)
```


## Generate volcano-plots. Figure 1B, Figure S2
```{r, echo=FALSE, warning=FALSE}
library(ggrepel)

  ggplot(data=results_sig, aes(x=Log_FC, y=Log10_p, col=Abundance, label=siglabel)) + 
   geom_point(alpha = 0.7, size=1, stroke=0) + 
   #geom_text_repel(size=2.0, force=3) +
   theme(text=element_text(size=8,  family="Arial"),
        panel.background = element_rect(fill = "white", colour = "grey50"),
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        axis.text.x = element_text(color = "black", size=5, angle = 90, vjust= 0.5), axis.text.y = element_text(color = "black", size=8),
        axis.title.x =element_text(color = "black", size=8), axis.title.y =element_text(color = "black", size=8),
         legend.position = "none")+
   geom_vline(xintercept=c(-0.6, 0.6), col="#3F5151", linetype = "dotted") +
   geom_hline(yintercept=-log10(0.05), col="#3F5151", linetype = "dotted") +
   scale_x_continuous(name="Log2 (Fold Change)",limits=c(-12,12), breaks=seq(-12, 12, 3))+
   scale_y_continuous(name="-Log10 (p-value)",limits=c(0,7), breaks=seq(0, 7, 1))+
   scale_color_manual(values=c("orangered1","steelblue2", "grey70")) +
   facet_wrap(~Condition, nrow =1, labeller = as_labeller(c("MD_P"="MD (P)", "Sn_SW"="Sn","Sn_P"="Sn (P)","Gi_SW"="Gi", 
                                                            "Gi_P"="Gi (P)","Sn_Gi_SW"="Sn+Gi","Sn_Gi_P"="Sn+Gi (P)")))
 
 #  ggsave("Sn_All_volcano_facet_v2.pdf",device=cairo_pdf, width=6, height= 3, units = "in", dpi = 300) 
  
 ggplot(data=results_sig_1, aes(x=Log_FC, y=Log10_p, col=Abundance, label=siglabel)) + 
   geom_point(alpha = 0.7, size=1.0, stroke=0) + 
   #geom_text_repel(size=1.0, force=1, nudge_y = 0.5) +
   theme(panel.background = element_rect(fill = "white", colour = "grey50"),
         strip.background = element_rect(fill="white", size = 3),
         strip.text=element_text(color = "black", size=8),
         legend.position = "none", 
         axis.text.x = element_text(color = "black", size=6),
         axis.text.y = element_text(color = "black", size=6),
         axis.title.x = element_text(color = "black",size = 8, face = "bold"), 
         axis.title.y = element_text(color = "black",size = 8, face = "bold")) +
   geom_vline(xintercept=c(-0.6, 0.6), col="#3F5151", linetype = "dashed", size=0.25) +
   geom_hline(yintercept=-log10(0.05), col="#3F5151", linetype = "dashed", size =0.25) +
   geom_text(size=1.0, nudge_y = 0.25) +
   scale_x_continuous(name="Log2 (Fold Change)",limits=c(-10,10), breaks=seq(-10, 10, 5))+
   scale_y_continuous(name="-Log10 (p-Value)",limits=c(0,6), breaks=seq(0, 6, 1))+
   scale_color_manual(values=c("orangered1","steelblue2", "grey70")) +
   facet_grid(vars(Food), vars(Colonization))
 # ggsave("Sn_Gi_volcano_facet_4.pdf",device=cairo_pdf, width=6, height= 4, units = "in", dpi = 300) 
 
   ggplot(data=results_sig_sn, aes(x=Log_FC, y=Log10_p, col=Abundance, label=siglabel)) + 
   geom_point(alpha = 0.7, size=1.5, stroke=0) + 
   #geom_text_repel(size=1.0, force=1, nudge_y = 0.5) +
   theme(panel.background = element_rect(fill = "white", colour = "grey50"),
         strip.background = element_rect(fill="white", size = 3),
         strip.text=element_text(color = "black", size=8),
         legend.position = "none", 
         axis.text.x = element_text(color = "black", size=6),
         axis.text.y = element_text(color = "black", size=6),
         axis.title.x = element_text(color = "black",size = 8, face = "bold"), 
         axis.title.y = element_text(color = "black",size = 8, face = "bold")) +
   geom_vline(xintercept=c(-0.6, 0.6), col="#3F5151", linetype = "dashed", size=0.25) +
   geom_hline(yintercept=-log10(0.05), col="#3F5151", linetype = "dashed", size =0.25) +
   scale_x_continuous(name="Log2 (Fold Change)",limits=c(-10,10), breaks=seq(-10, 10, 5))+
   scale_y_continuous(name="-Log10 (p-Value)",limits=c(0,6), breaks=seq(0, 6, 1))+
   scale_color_manual(values=c("orangered1","steelblue2", "grey70")) +
   facet_wrap(~Condition, nrow =1, labeller = as_labeller(c("Sn_SW"="Sugar Water", "Sn_P"="Pollen")))
 #  ggsave("Sn_volcano_facet_F2.pdf",device=cairo_pdf, width=4, height= 3, units = "in", dpi = 300) 


   
```

##Calculate Z-Scores of metabolites
```{r, message = FALSE, warning=FALSE}

Results_NC_samples <- Results_NC %>%
  filter(Type != "PE")%>%
  group_by(Metabolite) %>%
  mutate(z_score = (Norm_resp - mean(Norm_resp, na.rm = TRUE))/sd(Norm_resp, na.rm = TRUE),
         LOD = (1-mean(Norm_resp, na.rm = TRUE))/sd(Norm_resp, na.rm = TRUE)) %>%
  ungroup(Metabolite)

Results_NC_samples$Condition <- factor(Results_NC_samples$Condition, 
                                     levels = c("MD_SW", "MD_P", "Sn_SW", "Sn_P", "Gi_SW", "Gi_P", "Sn_Gi_SW", "Sn_Gi_P"))
```


## Test for differentially abundant metabolites using mixed linear models
### Implement mixed linear model of lmm2met
#### Variables: Fixed (colonization and injection order), Random (batch) 

```{r, message=FALSE, warning=FALSE}
metabdf <- Results_NC_samples %>%
  select(Sample, Metabolite) %>%
  filter(Sample == "A12FF5")
metabdf$metabID<-1:nrow(metabdf)
metabdf$metabID = paste0('X', metabdf$metabID)
metabdf$metabID <- factor(metabdf$metabID)

metabdf <- metabdf %>%
  select(Metabolite, metabID)

metab_lmm <- Results_NC_samples %>%
  select(Sample,Bee_exp,Condition,Food,Metabolite,z_score) %>%
  mutate(Sn=case_when(str_detect(Condition,"Sn")~1,
                      TRUE ~ 0),
         Gi=case_when(str_detect(Condition,"Gi")~1,
                      TRUE ~ 0))%>%
  full_join(metabdf) %>%
  ungroup() %>%
  group_by(metabID) %>%
  select(Sample, Bee_exp, Sn, Gi, Food, metabID, z_score) %>%
  pivot_wider(names_from = metabID, values_from = z_score)%>%
  replace(is.na(.), 0)


metab_lmm$Food <- factor(metab_lmm$Food, levels = c("SW", "P"))
metab_lmm$Sn <- factor(metab_lmm$Sn)
metab_lmm$Gi <- factor(metab_lmm$Gi)
metab_lmm$Bee_exp <- factor(metab_lmm$Bee_exp)

library(lmm2met)
fitMet = fitLmm(fix= c('Food','Sn',"Gi"), random='(1|Bee_exp)', data=metab_lmm, start=6, end=238)

plot(fitMet, type='coeff')

plot(fitMet, type='residual') 

plot(fitMet, type='fixeff')

plot(fitMet, type='randeff') 

plot(fitMet, type='chisq')

getFixCoeff(fitMet, save=TRUE) 

list2env(fitMet,envir = .GlobalEnv)

fittedDat_long <- fittedDat %>%
  pivot_longer(cols =  c("X1":"X233"),
               names_to = "Metabolite",
               values_to = "FittedDat",
               values_drop_na = FALSE)
```
## Adjust p-values from previous chunk to account for multiple testing
## output.csv generated in previous chunk
```{r, message=FALSE, warning=FALSE}
lmm_output_csv <- read_csv("output.csv")
lmm_output <- lmm_output_csv %>% 
  rename(p_Diet = "Pr(Chi).Food", metabID = 1, p_Sn = "Pr(Chi).Sn",p_Gi = "Pr(Chi).Gi",Sn="Sn1",Gi="Gi1", Pollen="FoodP") %>%
  mutate(p_Diet_adj = p.adjust(p_Diet, method="BH",n = length(p_Diet)),
    Log10_p_Diet = -log10(p_Diet_adj),
    p_Sn_adj = p.adjust(p_Sn, method="BH",n = length(p_Sn)),
    Log10_p_Sn = -log10(p_Sn_adj),
    p_Gi_adj = p.adjust(p_Gi, method="BH",n = length(p_Gi)),
    Log10_p_Gi = -log10(p_Gi_adj)) %>%
  left_join(metabdf) %>%
  mutate(Species_Significance = case_when((p_Sn_adj < 0.05) & (p_Gi_adj >= 0.05)  ~ "Sn",
                                  (p_Sn_adj >= 0.05) & (p_Gi_adj < 0.05)  ~ "Gi",
                                  (p_Sn_adj < 0.05) & (p_Gi_adj < 0.05)  ~ "Both",
                                  TRUE ~ "NS"),
         Diet_Significance = case_when((p_Diet_adj < 0.05) ~ "Pollen",
                                       TRUE ~ "NS"))%>%
  unite("Significance", Species_Significance:Diet_Significance, remove = FALSE)

lmm_output$siglabel <- NA
lmm_output$Metabolite <- as.character(lmm_output$Metabolite)
lmm_output$siglabel[lmm_output$p_Sn <= .05] <- lmm_output$Metabolite[lmm_output$p_Sn <= .05]

lmm_output_sig <- lmm_output %>%
  filter(Species_Significance != "NS") 
```


## Generate fixed effects plot for Figure 1C
```{r, echo=FALSE, warning=FALSE}

ggplot(data=lmm_output_sig, aes(x=Sn, y=Gi, color=Pollen, label=siglabel)) + 
  geom_vline(xintercept=0, col="black", linetype = "solid") +
  geom_hline(yintercept=0, col="black", linetype = "solid") +
  geom_vline(xintercept=-.4, col="grey", linetype = "dashed", size=0.25) +
  geom_vline(xintercept=.4, col="grey", linetype = "dashed", size=0.25) +
  geom_hline(yintercept=-.4, col="grey", linetype = "dashed", size=0.25) +
  geom_hline(yintercept=.4, col="grey", linetype = "dashed", size=0.25)+
  geom_point(size=1) +
  theme(panel.background = element_rect(fill = "white", colour = "black"),
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        strip.background = element_rect(fill="white",colour = "black"),
        legend.position = "right",
        axis.text.x = element_text(color = "black", size=6),
        axis.text.y = element_text(color = "black", size=6),
        axis.title.x = element_text(color = "black", size=8), 
        axis.title.y = element_text(color = "black", size=8),
        legend.text=element_text(color = "black", size=6),
        legend.title=element_text(color = "black", size=6))+
  geom_text_repel(size = 2, max.overlaps = Inf, color="slategrey")+
    binned_scale(aesthetics = "color",
               scale_name = "stepsn", 
               palette = function(x) c( "#313695","#4575B4","#74ADD1","#ABD9E9","grey",
                                        "#FDAE61","#F46D43","#D73027","#A50026"),
               breaks = c(-1.25, -1, -.75, -.5, .5, .75, 1, 1.25),
               limits = c(-1.5, 1.5),
               show.limits = TRUE, 
               guide = "colorsteps")+
  scale_x_continuous(name="Effect Size (S. alvi)",limits=c(-1.25,1.25), 
                     breaks=seq(-1.25,1.25, 0.25), labels = scales::comma)+
  scale_y_continuous(name="Effect Size (Gilliamella)",limits=c(-1.25,1.25), 
                     breaks=seq(-1.25,1.25, 0.25), labels =  scales::comma)
#ggsave("Effect_size_1C.pdf",device=cairo_pdf, width=5, height= 3, units = "in", dpi = 300) 

```


## Generate plot for Figure S3
```{r, echo=FALSE, warning=FALSE}   
ggplot(subset(Results_NC_samples, Metabolite %in% 
                c("Fumarate", "Acetate","alpha-Ketoglutarate", "3-Hydroxy-3-methylglutarate","Lactate", "Pyruvate","Citrate", "Glycerate","Unknown_16_22","Unknown_15_15", "Unknown_12_74", "Benzoic acid", "Succinate","Dihydroxynaphthalene", "Anthranilate", "Glycine", "Caffeic_acid", "Urea", "2-Ketoisocaproate", "Malate", "Aspartate")), 
       mapping = aes(x = Condition, y = z_score, fill = Condition)) +
  geom_boxplot(na.rm = TRUE, alpha = 1, outlier.shape = NA, lwd=0.25) +
  geom_jitter(width = 0.15, height = 0, size=0.75, alpha=0.8, stroke=0) +
  scale_fill_manual(values=c("grey", "slategrey","plum1", "plum4", 
                             "skyblue","royalblue", "palegreen", "forestgreen")) +
  stat_n_text(na.rm = TRUE, size = 1.5, y.pos = -1.8 ) +
  geom_hline(aes(yintercept=LOD), col="#3F5151", linetype = "dotted") +
  theme(panel.background = element_rect(fill = "white", colour = "black"),
        panel.border = element_rect(colour = "grey", fill=NA, size=1),
        strip.background = element_rect(fill="white",colour = "black"),
        legend.position = "none",
        axis.text.x = element_blank(),
        axis.text.y = element_text(color = "black", size=8),
        axis.title.x = element_blank(), 
        axis.title.y = element_text(color = "black", size=10),
        strip.text.x = element_text(size = 6)) +
  scale_y_continuous(limits=c(-1.8,6), breaks=seq(-2, 6, 1)) +
  #scale_x_discrete(breaks=c('MD_SW','Sn_SW','Gi_SW', 'Sn_Gi_SW'), labels=c('MD','Sn','Gi', 'Sn+Gi'))+
  stat_compare_means(label = "p.signif", method = "wilcox.test", 
                     ref.group = "MD_SW", hide.ns = TRUE, na.rm = TRUE,
                     label.y = 5.0, color = "grey") +
    stat_compare_means(label = "p.signif", method = "wilcox.test", 
                       ref.group = "MD_P", hide.ns = TRUE, na.rm = TRUE,
                     label.y = 5.25, color = "slategrey") +
  facet_wrap(~Metabolite,nrow=4)
  # ggsave("Fig_S3_z_score_box.pdf",device=cairo_pdf, width=9, height= 7, units = "in", dpi = 300)    
```

## Fig 2A: Compare Metabolite Abundances in Pollen Extract vs Bee Guts
### Bees consumed ~21 mg Pollen
### Gut samples contain 4.3X more pollen than in same volume of Extract 
```{r, message = FALSE, warning=FALSE} 
Results_PE <- Results_NC %>%
  filter(Condition == "PE" | Condition == "MD_SW" | Condition == "MD_P")%>%
  mutate(Norm_resp_pe = case_when(Type == "PE" ~ 4.3*Norm_resp,
                                  TRUE ~ Norm_resp)) %>%
  select(Metabolite,Sample,Condition,Norm_resp_pe)

Group_mean <-Results_PE %>%
  group_by(Metabolite,Condition)%>%
  summarise(Avg = mean(Norm_resp_pe),
            Sd = sd(Norm_resp_pe))%>%
  pivot_wider(names_from="Condition", values_from = c("Avg", "Sd"))%>%
  mutate(Rel_MD_SW = Avg_MD_SW / Avg_MD_P,
         Rel_PE = Avg_PE / Avg_MD_P,
         stdev_MD_SW = Rel_MD_SW*sqrt((Sd_MD_SW/Avg_MD_SW)^2+(Sd_MD_P/Avg_MD_P)^2),
         stdev_PE = Rel_PE*sqrt((Sd_PE/Avg_PE)^2+(Sd_MD_P/Avg_MD_P)^2))%>%
  pivot_longer(
   cols = starts_with("Rel_"),
   names_to = "Condition",
   names_prefix = "Rel_",
   values_to = "Frac_abundance",
   values_drop_na = TRUE) 


Group_mean_plot <- Group_mean %>%
    filter(Metabolite %in% c("3-Hydroxy-3-methylglutarate", "alpha-Ketoglutarate",
                           "Aspartate","Citrate","Fumarate","Glutamate","Glycerate",
                           "Kynurenine","Lactate","Malate","Putrescine","Pyruvate", "Alanine", "Methionine",                                 "Phenylalanine", "Tyrosine", "Valine", "Caffeic_acid"))%>%
  mutate(Condition= fct_relevel(Condition, "PE", "MD_SW"),
         Metabolite = fct_recode(Metabolite, "3HMG" = "3-Hydroxy-3-methylglutarate", "a-Ketoglutarate" = "alpha-Ketoglutarate"))

ggplot(Group_mean_plot, mapping = aes(x = fct_reorder(Metabolite, Frac_abundance), y = Frac_abundance, fill = Condition)) +
  geom_col(size=1, alpha=1, stroke=0) +
  geom_hline(yintercept=1, color="black",size=0.5)+
  scale_fill_manual(values=c("grey60", "grey30")) +
   theme(panel.border = element_rect(colour = "grey", fill=NA, size=1),
        panel.background = element_rect(fill = "white", colour = "grey50"), 
        axis.text.x = element_text(color = "black", size=6, angle=45, hjust=0.95, vjust=0.95), 
        axis.text.y = element_text(color = "black", size=6),
        axis.title.x =element_text(color = "black", size=8),
        axis.title.y =element_text(color = "black", size=8),
        legend.text = element_text(color = "black", size=8), 
        legend.title = element_text(color = "black", size=8),
        legend.position = c(0.7, 0.75),
        legend.key.size = unit(0.15, 'cm'),
        plot.title = element_text(hjust = 0.5, size=8))+
  scale_y_continuous(name="Fractional Abundance vs Pollen Filled Guts",limits=c(0,9), breaks=seq(0, 9, 0.5))+
  scale_x_discrete(name="Metabolite")
#ggsave("Fig2a_Frac_abundance.pdf" ,device=cairo_pdf, width=5, height= 3, units = "in", dpi = 300)



```

